preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-07 Manuscript/manuscript.Rmd
Removing most ???

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools Arrowhead [@durand2016juicebox] and Peakachu [???change to lowercase] [@salameh2019supervised] respectively. Arrowhead [???change to lowercase] TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly).

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base resolution.

We use called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publicly available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the Arrowhead pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that Arrowhead data constitute the main results. For MD]

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). [???need concluding sentence]

@@

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E) [???should I mention jaccard or scrap it? significantly different for arrowhead but not peakachu].

@@

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting boundary regions.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 4-5 [??? Merge them into one Excel with two sheets - same message]!!!will do when ready). Our aim was to have a balanced metric that was sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it? for MD]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent].  [??? So what?]. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. [??? So what?]

@@

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and a fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to high proportion of TAD boundaries being undetected by *preciseTAD* despite being detected by TAD callers [??? What is the false positive rate? We may be missing TADs that are distinct from the canonical CTCF boundaries. TBD; !!!discuss in meeting]. Our future work will include defining separate models trained on different TAD boundary types, and predicting precise locations for each of them.

>>>>

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. Arrowhead TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly).

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict the genomic coordinates of TAD and loop boundaries at base resolution.

We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1).

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). These results suggest that a classifier trained on the data from one cell line performs well when using data from another cell line.

@@

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). <!--Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two domains are inherently conjoined by only one predicted boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] -->Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E).

@@

Our machine learning framework yielded several interesting observations. First, we demonstrated that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique, an aspect of boundary prediction that was unaddressed in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa]. We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent; @sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. <!--Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. -->This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting domain boundaries.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric that was sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it?]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike Arrowhead, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* TAD and loop boundaries were enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143, supporting recent observations that TADs and loops are dynamic structures generated by similar mechanisms [@hansen2018recent]. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, a known feature among the 3D architecture of the human genome, further highlighting their biological significance [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].

@@

Another limitation of *preciseTAD* is that it does not distinguish boundary types. The hierarchical nature of TAD boundaries [@weinreb2016identification; @fraser2015hierarchical; @Cresswell:2020ab] is not considered by *preciseTAD* due to the lack of gold standard of TAD hierarchy. Recent research distinguishes CTCF-associated boundaries, CTCF-negative YY1-enriched boundaries, CTCF- and YY1- depleted promoter boundaries, and a fourth class of weak boundaries largely depleted of all three features [@Krietenstein:2020aa]. Furthermore, actively transcribed regions can serve as TAD boundaries themselves, independently of CTCF binding [@Harrold:2020aa]. This may lead to some TAD boundaries being undetected by *preciseTAD* despite being detected by TAD callers. Our future work will include defining separate models trained on different TAD boundary types, and predicting precise locations for each of them.
